---
title: "Part 1: Tweedie fit and the single-index test"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

Understat shot data, EPL 2014/15 to 2024/25. A team's match npxG is the
sum of its non-penalty shot xG values, which factors exactly into volume
and mean severity:

$$S = \sum_{i=1}^{N} X_i = N \cdot \bar X$$

The compound Poisson-Gamma (Tweedie) model is the usual choice for $S$.
With constant dispersion, its index $p \in (1,2)$ simultaneously fixes
the variance function, the elasticity split between $N$ and $\bar X$,
and the Gamma shape of individual shots. Part 1 estimates $p$ from all
three and asks whether they agree. Part 2 (`nbg_fit.Rmd`) fits the two
components separately.

Mean structure throughout, in both parts:

$$\log \mu_{i} = \text{season}_{s(i)} + \beta\,\text{home}_i + \text{ATT}_{a(i)} + \text{DEF}_{d(i)}$$

with team-season attack and defence effects, sum-to-zero within season.
No form covariates.

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r config}
library(dplyr)
library(ggplot2)
library(Matrix)
library(mcglm)
source("data_pipeline.R")

LEAGUE  <- "EPL"
SEASONS <- 2014:2024
set.seed(2026)
dir.create("output", showWarnings = FALSE)
```

## 1. Data

Penalties and own goals are removed from the fitted target, the shot
counts and the shot-quality distribution alike, so all three estimates
of $p$ use the same shots. Zero-shot sides are kept at xG = 0.

```{r data}
xg_data <- prepare_xg_data(LEAGUE, SEASONS, exclude_penalties = TRUE)
shots   <- xg_data$shots
stacked <- xg_data$stacked

cat("Matches:", nrow(xg_data$matches), "\n")
cat("Team-match rows:", nrow(stacked), "\n")
cat("Rows with zero shots:", sum(stacked$n_shots == 0), "\n")
cat("Shots:", nrow(shots), "\n")
cat("Mean shots per team-match:", round(mean(stacked$n_shots), 2), "\n")
cat("Mean npxG per team-match:", round(mean(stacked$xg), 3), "\n")
```

## 2. Team-season volume vs. quality

Descriptive, before any model: do teams reach their xG through volume or
severity? Grey curves are constant npxG per match
($N \cdot \bar X = k$), so movement along a curve trades one for the
other.

```{r tactical-plot}
team_seasons <- stacked |>
  group_by(att_name, season) |>
  summarise(volume = mean(n_shots),
            quality = sum(xg) / sum(n_shots),
            xg_pm = mean(xg), .groups = "drop") |>
  mutate(label = paste(att_name, season))

to_label <- bind_rows(slice_max(team_seasons, xg_pm, n = 5),
                      slice_min(team_seasons, xg_pm, n = 3))

iso_curves <- lapply(c(0.8, 1.2, 1.6, 2.0, 2.4), function(k)
  geom_function(fun = function(v) k / v, colour = "grey70"))

ggplot(team_seasons, aes(volume, quality)) +
  iso_curves +
  geom_point() +
  geom_text(data = to_label, aes(label = label), size = 3, vjust = -0.8) +
  coord_cartesian(ylim = range(team_seasons$quality)) +
  labs(x = "Shots per match", y = "npxG per shot") +
  theme_minimal()

cat("SD of log volume: ", round(sd(log(team_seasons$volume)), 3), "\n")
cat("SD of log quality:", round(sd(log(team_seasons$quality)), 3), "\n")
```

## 3. Tweedie fit

$\mathrm{Var}(Y) = \phi\,\mu^{p}$ with $p$ estimated, fitted in `mcglm`
with two covariance components: an identity term and a shared match
term, so the two sides of a fixture are allowed to be correlated.

Rank check first: mcglm reports a rank-deficient design with a
misleading error, so fail here with the aliased columns named instead.

```{r rank-check}
rhs <- model_rhs(stacked)
form_tweedie <- stats::reformulate(rhs, response = "xg")

X_check <- model.matrix(form_tweedie, data = stacked)
qrX <- qr(X_check)
if (qrX$rank < ncol(X_check)) {
  stop("Design is rank deficient. Aliased: ",
       paste(head(colnames(X_check)[qrX$pivot[(qrX$rank + 1):ncol(X_check)]], 10), collapse = ", "))
}
cat("Regression parameters:", ncol(X_check), "on", nrow(stacked), "rows, full rank\n")
rm(X_check, qrX)
```

```{r tweedie-fit}
Z_id    <- mc_id(stacked)
Z_match <- list(tcrossprod(Matrix::sparse.model.matrix(~ 0 + match_f, data = stacked)))

fit_tweedie <- mcglm(
  linear_pred = c(form_tweedie), matrix_pred = list(c(Z_id, Z_match)),
  link = "log", variance = "tweedie", power_fixed = FALSE,
  data = stacked,
  control_algorithm = list(tuning = 0.5, max_iter = 100, correct = TRUE, verbose = FALSE)
)

coefs <- coef(fit_tweedie, std.error = TRUE)
p_hat <- coefs$Estimates[coefs$Type == "power"]
p_se  <- coefs$Std.error[coefs$Type == "power"]
stacked$mu_tweedie <- as.numeric(fitted(fit_tweedie))

coefs[coefs$Type != "beta", ]
cat("Implied split at p_hat: volume", round(2 - p_hat, 3),
    " quality", round(p_hat - 1, 3), "\n")
```

Home advantage as a multiplier on xG, and the within-match correlation.
The covariance has tau11 on the diagonal plus tau12 for every pair of
rows in the same match, so the implied correlation between the two sides
is tau12 / (tau11 + tau12).

```{r home-and-match}
home_i  <- match("home", fit_tweedie$beta_names[[1]])
home_b  <- coef(fit_tweedie, std.error = TRUE, type = "beta")[home_i, ]
cat("Home multiplier:", round(exp(home_b$Estimates), 3),
    " 95% CI:", round(exp(home_b$Estimates + c(-1.96, 1.96) * home_b$Std.error), 3), "\n")

tau <- coefs$Estimates[coefs$Type == "tau"]
cat("Implied within-match correlation:", round(tau[2] / (tau[1] + tau[2]), 3), "\n")
cat("Raw correlation of home and away npxG:",
    round(with(xg_data$matches, cor(home_xg, away_xg)), 3), "\n")
```

## 4. Test 1: volume/quality elasticities

Under the compound model with constant dispersion,
$E[N] \propto \mu^{2-p}$ and $E[\bar X] \propto \mu^{p-1}$. Regressing
each component on $\log\hat\mu$ therefore recovers the split the fitted
$p$ implies.

Since $S = N \bar X$, the two slopes sum to about 1 whenever $\hat\mu$
is calibrated, so the sum is a check on the fit rather than a test of
the Tweedie. The test is the split. The SEs below treat $\log\hat\mu$ as
fixed and ignore the within-match correlation.

```{r elasticity}
el_dat <- stacked |>
  filter(mu_tweedie > 0) |>
  mutate(log_mu = log(mu_tweedie))

glm_volume  <- glm(n_shots ~ log_mu, family = quasipoisson(link = "log"), data = el_dat)
glm_quality <- glm(xg_mean_shot ~ log_mu, family = Gamma(link = "log"),
                   weights = n_shots, data = filter(el_dat, n_shots > 0, xg_mean_shot > 0))

e_vol  <- summary(glm_volume)$coefficients["log_mu", 1:2]
e_qual <- summary(glm_quality)$coefficients["log_mu", 1:2]

elasticity <- data.frame(
  component = c("volume", "quality"),
  observed  = c(e_vol[1], e_qual[1]),
  se        = c(e_vol[2], e_qual[2]),
  tweedie   = c(2 - p_hat, p_hat - 1)
)
elasticity$z <- (elasticity$observed - elasticity$tweedie) / sqrt(elasticity$se^2 + p_se^2)
elasticity

cat("Sum of elasticities:", round(e_vol[1] + e_qual[1], 3), "\n")
cat("Quasi-Poisson dispersion (shots on Tweedie mean):",
    round(summary(glm_volume)$dispersion, 3), "\n")
```

## 5. Test 2: shot-quality shape

The compound model makes each shot Gamma with shape
$\alpha = (2-p)/(p-1)$, equivalently

$$p = \frac{\alpha + 2}{\alpha + 1}$$

Method-of-moments shape ($\hat\alpha = \bar x^2 / s^2$), pooled and
within team-season (each shot divided by its team-season mean, removing
between-team differences in average severity). Intervals from a
bootstrap that resamples whole matches.

```{r shape}
shot_dat <- shots |>
  mutate(team = ifelse(side == "h", home_team, away_team)) |>
  group_by(team, season) |>
  mutate(ratio = xG / mean(xG)) |>
  ungroup()

alpha_pooled <- gamma_shape_mom(shot_dat$xG)
alpha_within <- 1 / var(shot_dat$ratio)

by_match <- shot_dat |>
  group_by(match_id) |>
  summarise(n = n(), sx = sum(xG), sx2 = sum(xG^2),
            sr = sum(ratio), sr2 = sum(ratio^2), .groups = "drop")

mom_alpha <- function(n, s, s2) {
  N <- sum(n)
  m <- sum(s) / N
  v <- (sum(s2) - N * m^2) / (N - 1)
  m^2 / v
}

boot <- replicate(200, {
  b <- by_match[sample.int(nrow(by_match), replace = TRUE), ]
  c(pooled = mom_alpha(b$n, b$sx, b$sx2), within = mom_alpha(b$n, b$sr, b$sr2))
})

shape <- data.frame(
  estimate = c("pooled", "within team-season"),
  alpha    = c(alpha_pooled, alpha_within),
  alpha_lo = apply(boot, 1, quantile, 0.025),
  alpha_hi = apply(boot, 1, quantile, 0.975)
)
shape$p    <- p_from_shape(shape$alpha)
shape$p_lo <- p_from_shape(shape$alpha_hi)
shape$p_hi <- p_from_shape(shape$alpha_lo)
shape

cat("Shape implied by the observed split:", round(e_vol[1] / (1 - e_vol[1]), 3), "\n")
cat("Shot-shape p vs fit, in SEs of p:", round((shape$p[2] - p_hat) / p_se, 1), "\n")
```

## 6. All estimates of p

```{r p-compare}
p_tbl <- data.frame(
  source = c("Tweedie fit", "Volume elasticity", "Quality elasticity",
             "Shape, pooled", "Shape, within team-season"),
  p  = c(p_hat, 2 - e_vol[1], 1 + e_qual[1], shape$p),
  lo = c(p_hat - 1.96 * p_se, 2 - e_vol[1] - 1.96 * e_vol[2],
         1 + e_qual[1] - 1.96 * e_qual[2], shape$p_lo),
  hi = c(p_hat + 1.96 * p_se, 2 - e_vol[1] + 1.96 * e_vol[2],
         1 + e_qual[1] + 1.96 * e_qual[2], shape$p_hi)
)
p_tbl$source <- factor(p_tbl$source, levels = rev(p_tbl$source))
p_tbl

ggplot(p_tbl, aes(p, source)) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2) +
  geom_point() +
  xlim(1, 2) +
  labs(x = "Tweedie index p", y = NULL) +
  theme_minimal()
```

The fit and both elasticities agree near $p \approx 1.3$; the shape
estimates sit near 1.66, several SEs of $\hat p$ above the fit, with
tight intervals. No single index satisfies both the team-level split and
the shot-level skew. Which of the compound model's assumptions fails
(Poisson counts, Gamma severity, independence) cannot be read off here,
which motivates Part 2.

```{r save}
saveRDS(stacked |> select(match_id, home, att_name, def_name, season, xg, n_shots, mu_tweedie),
        file.path("output", "tweedie_fitted.rds"))
```
