---
title: "Part 2: Negative Binomial-Gamma decomposition"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

Part 1 showed that one Tweedie index cannot match both the
volume/severity split and the skew of individual shots. Here the two
components are fitted separately, with the same mean structure (season,
home, team-season attack and defence), so each can carry its own
distribution and be checked on its own:

$$N \sim \text{NegBin}(\mu_N, \theta), \qquad \bar X \sim \text{Gamma}(\mu_X, \phi/N),
\qquad \widehat{S} = \hat\mu_N \cdot \hat\mu_X$$

Since both components are log-linear in the same covariates, the product
is too, so the recombined mean is in the same family as the Tweedie's.
The decomposition is for diagnosis and interpretation, not for better
point prediction; Section 5 confirms nothing is lost.

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r config}
library(dplyr)
library(ggplot2)
library(MASS, exclude = "select")
source("data_pipeline.R")

LEAGUE  <- "EPL"
SEASONS <- 2014:2024
TWEEDIE_FILE <- file.path("output", "tweedie_fitted.rds")  # written by tweedie_fit.Rmd
```

## 1. Data

Same build as Part 1.

```{r data}
xg_data <- prepare_xg_data(LEAGUE, SEASONS, exclude_penalties = TRUE)
shots   <- xg_data$shots
stacked <- xg_data$stacked
rhs     <- model_rhs(stacked)
C_team  <- attr(stacked, "contrasts_team")   # saved now; later joins drop attributes

quality_rows <- stacked$n_shots > 0 & stacked$xg_mean_shot > 0
cat("Team-match rows:", nrow(stacked), "\n")
cat("Rows used for quality model:", sum(quality_rows), "\n")
```

## 2. Volume

Is the Poisson assumption tenable?
$\mathrm{Var}(N) = \mu + \mu^2/\theta$, so the variance-to-mean ratio is
$1 + \mu/\theta$ at volume $\mu$.

```{r volume}
form_volume <- stats::reformulate(rhs, response = "n_shots")

fit_pois <- glm(form_volume, family = poisson, data = stacked)
fit_qp   <- glm(form_volume, family = quasipoisson, data = stacked)
fit_nb   <- glm.nb(form_volume, data = stacked)

mean_shots <- mean(stacked$n_shots)
qp_disp    <- summary(fit_qp)$dispersion
theta      <- fit_nb$theta

cat("Quasi-Poisson dispersion:", round(qp_disp, 3), "\n")
cat("NB theta:", round(theta, 2), " SE:", round(fit_nb$SE.theta, 2), "\n")
cat("NB var/mean at average volume:", round(1 + mean_shots / theta, 3), "\n")
cat("AIC Poisson:", round(AIC(fit_pois), 1), " AIC NB:", round(AIC(fit_nb), 1), "\n")
cat("NB residual deviance / df:", round(deviance(fit_nb) / df.residual(fit_nb), 3), "\n")
```

Observed shot-count distribution vs. expected counts summed over all
team-matches.

```{r volume-plot}
k <- 0:40
count_fit <- data.frame(
  shots    = k,
  observed = as.numeric(table(factor(stacked$n_shots, levels = k))),
  poisson  = sapply(k, function(j) sum(dpois(j, fitted(fit_pois)))),
  negbin   = sapply(k, function(j) sum(dnbinom(j, size = theta, mu = fitted(fit_nb))))
)

ggplot(count_fit, aes(shots)) +
  geom_col(aes(y = observed), fill = "grey75") +
  geom_line(aes(y = poisson, colour = "Poisson")) +
  geom_line(aes(y = negbin, colour = "Negative Binomial")) +
  labs(x = "Shots in the match", y = "Team-matches", colour = NULL) +
  theme_minimal()
```

## 3. Quality

### 3a. Gamma GLM dispersion

With shot-count weights, $\mathrm{Var}(\bar X) = \phi\,\mu^2 / N$, and
deviance / df estimates $\phi$ rather than testing fit against 1. Under
iid Gamma shots of shape $\alpha$, the benchmark is
$\phi \approx 1/\alpha$, so that is what the estimate is compared with
here.

```{r quality}
form_quality <- stats::reformulate(rhs, response = "xg_mean_shot")
fit_gamma <- glm(form_quality, family = Gamma(link = "log"), weights = n_shots,
                 data = stacked[quality_rows, ])

ratio <- shots |>
  mutate(team = ifelse(side == "h", home_team, away_team)) |>
  group_by(team, season) |>
  mutate(ratio = xG / mean(xG)) |>
  pull(ratio)
alpha_within <- 1 / var(ratio)

gamma_disp <- summary(fit_gamma)$dispersion
cat("Gamma GLM dispersion:", round(gamma_disp, 3), "\n")
cat("Residual deviance / df:", round(deviance(fit_gamma) / df.residual(fit_gamma), 3), "\n")
cat("Expected under iid Gamma shots (1/alpha):", round(1 / alpha_within, 3), "\n")
cat("Ratio:", round(gamma_disp * alpha_within, 3), "\n")
```

### 3b. Shot-quality shape: single Gamma vs. two-component mixture

The real test of the Gamma is the shape of individual shot xG. Both fits
below are by maximum likelihood on positive non-penalty shots; the
alternative is a two-component normal mixture on $\log x$,

$$f(x) = \frac{1}{x}\left[w\,\varphi\!\left(\tfrac{\log x - m_1}{s_1}\right)\big/s_1 +
(1-w)\,\varphi\!\left(\tfrac{\log x - m_2}{s_2}\right)\big/s_2\right]$$

fitted by EM, as a stand-in for routine chances plus big chances. A
Gamma with shape below 1 is unimodal with its mode at zero and cannot
reproduce two modes.

```{r shape-fits}
x  <- shots$xG[shots$xG > 0]
lx <- log(x)

# Single Gamma MLE
gamma_nll <- function(par) -sum(dgamma(x, shape = exp(par[1]), rate = exp(par[2]), log = TRUE))
mom_alpha <- gamma_shape_mom(x)
g_opt   <- optim(c(log(mom_alpha), log(mom_alpha / mean(x))), gamma_nll, method = "BFGS")
g_shape <- exp(g_opt$par[1])
g_rate  <- exp(g_opt$par[2])

# Two-component normal mixture on log xG via EM
em_mix <- function(y, iters = 500, tol = 1e-8) {
  w <- 0.8
  m <- as.numeric(quantile(y, c(0.3, 0.9)))
  s <- rep(sd(y) / 2, 2)
  ll_old <- -Inf
  for (i in seq_len(iters)) {
    d1 <- w * dnorm(y, m[1], s[1])
    d2 <- (1 - w) * dnorm(y, m[2], s[2])
    r  <- d1 / (d1 + d2)
    w  <- mean(r)
    m  <- c(sum(r * y) / sum(r), sum((1 - r) * y) / sum(1 - r))
    s  <- sqrt(c(sum(r * (y - m[1])^2) / sum(r), sum((1 - r) * (y - m[2])^2) / sum(1 - r)))
    ll <- sum(log(d1 + d2))
    if (abs(ll - ll_old) < tol * abs(ll)) break
    ll_old <- ll
  }
  list(w = w, mean = m, sd = s)
}
mix <- em_mix(lx)

ll_gamma <- -g_opt$value
ll_mix   <- sum(log(mix$w * dnorm(lx, mix$mean[1], mix$sd[1]) +
                    (1 - mix$w) * dnorm(lx, mix$mean[2], mix$sd[2])) - lx)

cat("Gamma MLE shape:", round(g_shape, 3), " (MoM shape:", round(mom_alpha, 3), ")\n")
cat("AIC single Gamma:", round(-2 * ll_gamma + 4), "\n")
cat("AIC mixture:     ", round(-2 * ll_mix + 10), "\n")
cat("Mixture shares:", round(c(mix$w, 1 - mix$w), 3), "\n")
cat("Mixture component medians (xG):", round(exp(mix$mean), 3), "\n")
```

```{r shape-plot}
# Densities per log10 unit, to match a histogram on a log10 axis
grid <- data.frame(xg = 10^seq(log10(min(x)), 0, length.out = 500))
grid$gamma   <- dgamma(grid$xg, g_shape, g_rate) * grid$xg * log(10)
grid$mixture <- (mix$w * dnorm(log(grid$xg), mix$mean[1], mix$sd[1]) +
                 (1 - mix$w) * dnorm(log(grid$xg), mix$mean[2], mix$sd[2])) * log(10)

ggplot(data.frame(xg = x), aes(xg)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, fill = "grey75") +
  geom_line(data = grid, aes(y = gamma, colour = "Single Gamma")) +
  geom_line(data = grid, aes(y = mixture, colour = "Two-component mixture")) +
  scale_x_log10() +
  labs(x = "xG per shot (log scale)", y = "Density per log10 unit", colour = NULL) +
  theme_minimal()
```

## 4. What makes teams good: team-season volume and quality effects

Attack effects from the two component models, relative to each season's
average team and adjusted for opponents and venue. A volume multiplier
of 1.20 means 20% more shots than an average team would take in the same
fixture; a quality multiplier of 1.10 means 10% more xG per shot.

```{r team-effects}
vol_att  <- extract_effects(fit_nb, "ATT", C_team)
qual_att <- extract_effects(fit_gamma, "ATT", C_team)

team_eff <- inner_join(
  vol_att  |> dplyr::select(level, team, season, volume = multiplier, volume_se = std.error),
  qual_att |> dplyr::select(level, quality = multiplier, quality_se = std.error),
  by = "level"
) |>
  mutate(total = volume * quality)

cat("Correlation of volume and quality effects (log scale):",
    round(cor(log(team_eff$volume), log(team_eff$quality)), 3), "\n")
cat("SD of log volume effect: ", round(sd(log(team_eff$volume)), 3), "\n")
cat("SD of log quality effect:", round(sd(log(team_eff$quality)), 3), "\n")

# Strongest attacks overall, with how each gets there
team_eff |> arrange(desc(total)) |> head(10)

# Highest quality relative to volume
team_eff |> arrange(desc(quality)) |> head(10)
```

```{r team-effects-plot}
to_label <- bind_rows(slice_max(team_eff, total, n = 6),
                      slice_max(team_eff, quality, n = 3),
                      slice_min(team_eff, total, n = 3)) |>
  distinct(level, .keep_all = TRUE)

ggplot(team_eff, aes(volume, quality)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(alpha = 0.6) +
  geom_text(data = to_label, aes(label = paste(team, season)), size = 3, vjust = -0.8) +
  labs(x = "Volume effect (shot multiplier vs. season average)",
       y = "Quality effect (xG-per-shot multiplier vs. season average)") +
  theme_minimal()
```

## 5. Recombined predictions and calibration

$\widehat{S}_i = \hat\mu_{N,i}\,\hat\mu_{X,i}$, which is unbiased for
$E[S]$ only if $N$ and $\bar X$ are conditionally independent, the
compound model's third assumption and still untested. Calibration slope
is the slope of actual on predicted, with 1 meaning calibrated.
Regressing predicted on actual instead gives roughly $r^2$ for any
calibrated predictor, which is what produced the apparent attenuation in
the earlier version of this analysis; that slope is printed below for
reference.

```{r recombine}
stacked$pred_volume  <- predict(fit_nb, newdata = stacked, type = "response")
stacked$pred_quality <- predict(fit_gamma, newdata = stacked, type = "response")
stacked$pred_nbg     <- stacked$pred_volume * stacked$pred_quality

fit_metrics <- function(pred, actual) {
  cal <- coef(lm(actual ~ pred))
  c(rmse = sqrt(mean((pred - actual)^2)),
    r = cor(pred, actual),
    r2 = cor(pred, actual)^2,
    cal_slope = cal[["pred"]],
    cal_intercept = cal[["(Intercept)"]])
}

metrics <- rbind(nbg = fit_metrics(stacked$pred_nbg, stacked$xg))

have_tweedie <- file.exists(TWEEDIE_FILE)
if (have_tweedie) {
  tw <- readRDS(TWEEDIE_FILE) |> dplyr::select(match_id, home, mu_tweedie)
  stacked <- left_join(stacked, tw, by = c("match_id", "home"))
  metrics <- rbind(metrics, tweedie = fit_metrics(stacked$mu_tweedie, stacked$xg))
}
round(metrics, 3)

# For reference: slope of predicted on actual (the orientation of the old plot)
cat("Slope of predicted on actual:",
    round(coef(lm(pred_nbg ~ xg, data = stacked))[["xg"]], 3), "\n")
```

```{r calibration}
bins <- calibration_bins(stacked$pred_nbg, stacked$xg)
bins

ggplot(stacked, aes(pred_nbg, xg)) +
  geom_point(alpha = 0.1, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_errorbar(data = bins, aes(x = mean_pred, ymin = lo, ymax = hi),
                inherit.aes = FALSE, width = 0.04, colour = "red") +
  geom_point(data = bins, aes(mean_pred, mean_actual), colour = "red") +
  coord_cartesian(xlim = c(0, 3.5), ylim = c(0, 5)) +
  labs(x = "Predicted npxG", y = "Actual npxG") +
  theme_minimal()

if (have_tweedie) calibration_bins(stacked$mu_tweedie, stacked$xg)
```

## 6. Summary

- Volume: mildly overdispersed relative to Poisson; the Negative
  Binomial absorbs it.
- Severity: the Gamma's dispersion is close to the iid benchmark, but
  shot xG is bimodal and no single Gamma fits it. This is where the
  Tweedie's single-index constraint breaks.
- Team-season effects separate cleanly into a volume multiplier and a
  severity multiplier, which the aggregate Tweedie cannot provide.
- Recombined predictions match the Tweedie and are calibrated across the
  range, top decile included. The limit is explanatory power, not bias.

Next: a mixture or beta model for severity, a test of independence
between $N$ and $\bar X$, and player-level or game-state covariates.
